Your time series is (probably) too short

### 2017-07-07 writing/manuscript_draft.Rmd
added section to reference figure in supp mat

---
bibliography: White_bib.bib
fontsize: 12pt
geometry: margin=1in
header-includes: \usepackage{float} \usepackage{setspace}\doublespacing \usepackage{lineno}
  \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;} \linenumbers
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
  word_document: default
---

@@

Observational studies and population censuses have become a cornerstone of modern ecological research and conservation biology [@Magurran2010]. Longterm data is necessary to both understand population dynamics and to assess extinction risk of species. Even though many time series may now be considered "longterm", most are still short (i.e. less than X years). *I could have statistic on how long is considered "longterm" in papers*. Time series are typically short for a variety of reasons. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time (citations). There are, of course, notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others (citations).

@@

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized amodel specifically for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organim, 15 years was adequate to assess longterm trends in abundance. However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on single species.

In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address various questions. More specifically, we estimate the minimum time series length required to asssess longterm changes in abundance via simple linear regression. We examine 878 population time series to estimate this minimum time series length required.
<!-- In this paper, we evaluate the minimal time series length required to answer X questions. 1) How sensitivive are IUCN asssessments to time series lengths? 2) What minimum time series length is required to measure the entire time series trend? 3) How long does a time series need to be to measure the variance in population size (rare event questions)? We demonstrate our approach with X case studies and then show results for X species from the Blah Blah Database.  -->

@@

```{r,echo=F,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length. The horizontal, dashed line is our desired statistical power of 0.8. The vertical, dashed line is the minimum time required to achieve our desired statistical power. (c) Minimum time required for different values of the trend strength ($r$) and (d) minimum time required for different levels of population variability ($\\sigma$).\\label{fig:theoretical_approach}'}
setwd("~/Desktop/Research/time-series-project")
source("scripts/theoretical-models/calculate_power.R")
source("scripts/theoretical-models/simulate_pop.R")

@@

## Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]. The data is originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series. The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The database also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, longterm trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterian A2 to classify a species as vulnerable [@IUCN2012].

```{r,echo=F}
# Pull in results of biological characteristics comparison to min time series length required
setwd("~/Desktop/Research/time-series-project")
load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
```
For a subset of populations in the database ($n =$ `r nrow(LPI_pop_info)`), we were able to connect a database on biological characteristics (**cite database here**) like body size and generation length. All `r nrow(LPI_pop_info)` populations are within the Aves class. In the supplementary material, we explore how the minimum time required does not strongly correlate with any biological characteristics.

@@

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, each time series is a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999, @Giron-Nava2017]. Then we determined the $T_{min}$ required to be confident (i.e. have high enough statistical power) in an estimate of the slope coefficient from linear regression. Therefore, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the longterm, or ``true", time series also has a significant slope. This is a measure of statistical power (often denoted as $1- \beta$, where $\beta$ is the probability of a type II error). Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention (citation). <!--In Figure \ref{fig:empirical_approach_example}b, statistical power increases with more time sampled until our desired power of 0.8 at 16 years. Therefore, 16 years would be our estimate for $T_{min}$.-->

We also examine how minimum time series calculations changed depending on the question of interest. In the supplementary material, we show how to make similar calculations to determine the minimum time series required to estimate overall population growth rate, the variability in population size, and future extinction risk. Again, we took subsamples of time series data to determine $T_{min}$ for each of these questions.

@@

We determined the minimum time series length required to address a particular question of interest. More specifically we asked, what is the minimum time series length required to determine, via linear regression, the longterm trend in population abundance? The minimum time series length required is that which had high enough statistical power. We investigated this question through two approaches: simulation-based and empirical analyses. We briefly desribe results from the simulation approach and then discuss our emprirical approach.

@@

#only look at pops with significant longterm declines or increases
pop_info=pop_info[pop_info$trend_p_value<0.05,]
long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)

@@

We explored two different approaches to estimate the minimum time series length required to address a particular question of interest. We specifically asked, what is the minimum time series length required to determine longterm changes in abundance using linear regression? The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987, Johnson2015]. Our simulations support past work that longer time series are needed when the trend strength (i.e. rate of increase or decrease) is weak or when population variability is high [@Gerrodette1987].

@@

There has also been interest in determining the minimum time series length in the bird community [@Hatch2003, @Bart2004, @Seavy2007]. @Hatch2003 used seabird monitoring data to estimate the minimum time needed to achieve appropriate levels of power. He found that the minimum time required ranged between 11 and 69 years depending on species, trend strength, and study design. Similarly, @Bart2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations. They found that only about 42% of populations in the Breeding Bird Survey met their criteria. @Seavy2007 asked whether statistical power was even a useful framework for assessing longterm population trends. They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8). Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for precision based metrics ..

An important related question, is the optimal allocation of sampling effort in space versus time (citations). In a theoretical investigation of this question, @Rhodes2011 found that the optimal allocation of sampling depended strongly on temporal and spatial autocorrelation. If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa. Our work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (cite figure here). @Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a longterm census of the collared pika (\emph{Ochotona collaris}) found in the Yukon. They estimated longterm growth rates among three subpopulations over a 10-year period. They found that censuses less than 5 years may be misleading and that extrapolating from one population to another, even when nearby, may be difficult. Therefore, they argued more sampling should be conducted spatially than temporally [@Morrison2008].

@@

@Gerber1999 investigated the minimum time series required to estimate population growth of an endagered, but recovering, species. They used a longterm census to retroactively determine the minimum time series required to assess threat status. They found that only XX years were needed, XX years before the delisting decision was actually made. This highlights the importance of estimating the minimunm time series required as an earlier decision would have saved time and money [@Gerber1999].
-->

@@

We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required. We show that to identify longterm changes in abundance, 10-15 years of continuos monitoring are required (Fig. \ref{fig:min_time_dist}). In line with theoretical predictions, we also show that $T_{min}$ is strongly correlated with the overall trend in abundance, variability in abundance, and the temporal autocorrelation (Fig. \ref{fig:correlates}). Our work implies that for many populations, census data less than 10-15 years are probably not reliable. This stresses the importance of longterm monitoring programs. <!--(where do IUCN recommendations come from?)--> From both a scientific and management perspective estimates of $T_{min}$ are important. If a time series is too short, we are likely to make incorrect conclusions about longterm trends. In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999]. Future work can examine other species, with a wider range of life history characteristics. In addition, similar approaches can be used to determine the minimum time series length required to address other questions.

@@

stuff here

@@

>>>>

---
bibliography: White_bib.bib
fontsize: 12pt
geometry: margin=1in
header-includes: \usepackage{float} \usepackage{setspace}\doublespacing \usepackage{lineno}
  \usepackage[round]{natbib} \bibpunct[; ]{(}{)}{,}{a}{}{;} \linenumbers

output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
  word_document: default
---

@@

Observational studies and population censuses have become a cornerstone of modern ecological research and conservation biology [@Magurran2010]. long-term data is necessary to both understand population dynamics and to assess extinction risk of species. Even though many time series may now be considered "long-term", most are still short (i.e. less than X years). *I could have statistic on how long is considered "long-term" in papers*. Time series are typically short for a variety of reasons. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time (citations). There are, of course, notable exceptions including the Isle Royale study, Iowa corn study, cod, lynx-hair, and many others (citations).

@@

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized amodel specifically for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organim, 15 years was adequate to assess long-term trends in abundance. However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on single species.

In this paper, we use both simulations and a database of empirical time series to determine the minimum number of years required to address various questions. More specifically, we estimate the minimum time series length required to asssess long-term changes in abundance via simple linear regression. We examine 878 population time series to estimate this minimum time series length required.
<!-- In this paper, we evaluate the minimal time series length required to answer X questions. 1) How sensitivive are IUCN asssessments to time series lengths? 2) What minimum time series length is required to measure the entire time series trend? 3) How long does a time series need to be to measure the variance in population size (rare event questions)? We demonstrate our approach with X case studies and then show results for X species from the Blah Blah Database.  -->

@@

```{r,echo=F,eval=F,fig.cap='(a) example of simulated time series for 40 time periods and (b) statistical power versus the simulated time series length. The horizontal, dashed line is our desired statistical power of 0.8. The vertical, dashed line is the minimum time required to achieve our desired statistical power. (c) Minimum time required for different values of the trend strength ($r$) and (d) minimum time required for different levels of population variability ($\\sigma$).\\label{fig:theoretical_approach}'}
setwd("~/Desktop/Research/time-series-project")
source("scripts/theoretical-models/calculate_power.R")
source("scripts/theoretical-models/simulate_pop.R")

@@

## Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]. The data is originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series. The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The database also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterian A2 to classify a species as vulnerable [@IUCN2012].

```{r,echo=F}
# Pull in results of biological characteristics comparison to min time series length required
setwd("~/Desktop/Research/time-series-project")
load('cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
```
For a subset of populations in the database ($n =$ `r nrow(LPI_pop_info)`), we were able to connect a database on biological characteristics [@Myhrvold2015] like body size and generation length. All `r nrow(LPI_pop_info)` populations are within the Aves class. In the supplementary material, we explore how the minimum time required does not strongly correlate with any biological characteristics.

@@

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, each time series is a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999, @Giron-Nava2017]. Then we determined the $T_{min}$ required to be confident (i.e. have high enough statistical power) in an estimate of the slope coefficient from linear regression. Therefore, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the long-term, or ``true", time series also has a significant slope. This is a measure of statistical power (often denoted as $1- \beta$, where $\beta$ is the probability of a type II error). Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention (citation). <!--In Figure \ref{fig:empirical_approach_example}b, statistical power increases with more time sampled until our desired power of 0.8 at 16 years. Therefore, 16 years would be our estimate for $T_{min}$.-->

We also examine how minimum time series calculations changed depending on the question of interest. In the supplementary material, we show how to make similar calculations to determine the minimum time series required to estimate overall population growth rate (Fig. \ref{fig:growth_rate}), the variability in population size, and future extinction risk. Again, we took subsamples of time series data to determine $T_{min}$ for each of these questions.

@@

We determined the minimum time series length required to address a particular question of interest. More specifically we asked, what is the minimum time series length required to determine, via linear regression, the long-term trend in population abundance? The minimum time series length required is that which had high enough statistical power. We investigated this question through two approaches: simulation-based and empirical analyses. We briefly desribe results from the simulation approach and then discuss our emprirical approach.

@@

#only look at pops with significant long-term declines or increases
#pop_info=pop_info[pop_info$trend_p_value<0.05,]
#long_dat=subset(long_dat,long_dat$ID %in% pop_info$ID)
#long_time_names=subset(long_time_names,long_time_names %in% pop_info$ID)

@@

We explored two different approaches to estimate the minimum time series length required to address a particular question of interest. We specifically asked, what is the minimum time series length required to determine long-term changes in abundance using linear regression? The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987, Johnson2015]. Our simulations support past work that longer time series are needed when the trend strength (i.e. rate of increase or decrease) is weak or when population variability is high [@Gerrodette1987].

@@

There has also been interest in determining the minimum time series length in the bird community [@Hatch2003, @Bart2004, @Seavy2007]. @Hatch2003 used seabird monitoring data to estimate the minimum time needed to achieve appropriate levels of power. He found that the minimum time required ranged between 11 and 69 years depending on species, trend strength, and study design. Similarly, @Bart2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations. They found that only about 42% of populations in the Breeding Bird Survey met their criteria. @Seavy2007 asked whether statistical power was even a useful framework for assessing long-term population trends. They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8). Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for precision based metrics ..

An important related question, is the optimal allocation of sampling effort in space versus time (citations). In a theoretical investigation of this question, @Rhodes2011 found that the optimal allocation of sampling depended strongly on temporal and spatial autocorrelation. If spatial population dynamics were highly correlated, then it was better to sample more temporally, and vice versa. Our work supports this idea as populations with strong temporal autocorrelation needed less years of sampling (cite figure here). @Morrison2008 also studied the optimal allocation of sampling effort in space versus time, but used emprical data from a long-term census of the collared pika (\emph{Ochotona collaris}) found in the Yukon. They estimated long-term growth rates among three subpopulations over a 10-year period. They found that censuses less than 5 years may be misleading and that extrapolating from one population to another, even when nearby, may be difficult. Therefore, they argued more sampling should be conducted spatially than temporally [@Morrison2008].

@@

@Gerber1999 investigated the minimum time series required to estimate population growth of an endagered, but recovering, species. They used a long-term census to retroactively determine the minimum time series required to assess threat status. They found that only XX years were needed, XX years before the delisting decision was actually made. This highlights the importance of estimating the minimunm time series required as an earlier decision would have saved time and money [@Gerber1999].
-->

@@

We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required. We show that to identify long-term changes in abundance, 10-15 years of continuos monitoring are required (Fig. \ref{fig:min_time_dist}). In line with theoretical predictions, we also show that $T_{min}$ is strongly correlated with the overall trend in abundance, variability in abundance, and the temporal autocorrelation (Fig. \ref{fig:correlates}). Our work implies that for many populations, census data less than 10-15 years are probably not reliable. This stresses the importance of long-term monitoring programs. <!--(where do IUCN recommendations come from?)--> From both a scientific and management perspective estimates of $T_{min}$ are important. If a time series is too short, we are likely to make incorrect conclusions about long-term trends. In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999]. Future work can examine other species, with a wider range of life history characteristics. In addition, similar approaches can be used to determine the minimum time series length required to address other questions.

@@

stuff here \ref{fig:empirical_approach_example}

\clearpage

@@

 \setcounter{figure}{0}
 \renewcommand{\thefigure}{S\arabic{figure}}

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{(a) Population size of Bigeye tuna (\emph{Thunnus obesus}) over
time. The line is the best fit line from linear regresssion. (b)
Statistical power for different subsets of the time series in panel
a.\label{fig:empirical_approach_example}}
\end{figure}

\begin{longtable}[]{@{}lrrrr@{}}
\caption{Output of generalized linear model to examine predictors of the
minimum time required for determing long-term population
trends.\label{table:model_output}}\tabularnewline
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endfirsthead
\toprule
& Estimate & Std. Error & z value &
Pr(\textgreater{}\textbar{}z\textbar{})\tabularnewline
\midrule
\endhead
(Intercept) & 3.7405513 & 0.0242742 & 154.09595 & 0\tabularnewline
abs\_overall\_trend & -104.0533454 & 3.0941097 & -33.62949 &
0\tabularnewline
autocorrelation & -0.4899533 & 0.0431769 & -11.34757 & 0\tabularnewline
variance & 18.7366110 & 0.6404458 & 29.25557 & 0\tabularnewline
\bottomrule
\end{longtable}

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{Mimimum time required versus (a) generation length (years), (b)
litter size (n), (c) adult body mass (grams), (d) maximum longevity
(years), (e) egg mass (grams), and (f) incubation (days). The lines in
each plot represent the best fit line from linear
regression.\label{fig:biological_correlates}}
\end{figure}
